---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
options(digits = 4, width = 120)
```

# dbc: Dictionary-based cleaning
<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

Tools for creating and applying dictionaries of value-replacement pairs, to
clean non-valid values of numeric, categorical, or date-type variables within a
dataset.

### Installation

Install from GitHub with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("epicentre-msf/dbc")
```

### Example usage

#### Example datasets and dictionaries

```{r}
library(dbc)
data(ll1)
data(clean_num1)
data(clean_categ1)
data(dict_categ1)

ll1 # example messy dataset
```


#### Check and clean numeric variables

##### 1. Produce an initial dictionary of non-valid numeric values

```{r}
dict_clean_numeric <- check_numeric(
  ll1,
  vars = c("age", "contacts"), # cols that should be numeric
  fn = as.integer              # values not coercible by `fn` are non-valid
)

dict_clean_numeric
```

##### 2. Manually review non-valid values and give appropriate replacements, or use keyword ".na" to indicate that the value has been reviewed and cannot be corrected.

Normally one would do this step in a spreadsheet but we'll do it in R here for simplicity.

```{r}
dict_clean_numeric$replacement <- c(".na", "39", "10", ".na")
```

##### 3. Use the cleaning dictionary to clean numeric variables

```{r}
clean_numeric(
  ll1,
  vars = c("age", "contacts"),
  dict_clean = dict_clean_numeric,
  fn = as.integer
)
```

##### 4. If the original dataset is updated, we can repeat the cleaning steps while retaining the corrections already made in the previous cleaning dictionary.

*Check for new non-valid numeric values, after incorporating previous cleaning*

```{r}
dict_clean_numeric_update <- check_numeric(
  ll2,                             # same as ll1 but with 3 additional entries
  vars = c("age", "contacts"),
  dict_clean = dict_clean_numeric, # incorporate previous cleaning before checking
  fn = as.integer,
  return_all = TRUE                # return original cleaning dict + new entries
)

dict_clean_numeric_update
```

*Manually specify replacement for new non-valid entry*

```{r}
dict_clean_numeric_update$replacement[5] <- "6"
```

*Apply updated cleaning dictionary to updated dataset*

```{r}
clean_numeric(
  ll2,
  vars = c("age", "contacts"),
  dict_clean = dict_clean_numeric_update,
  fn = as.integer
)
```


#### Check and clean categorical variables

##### 1. Produce an initial dictionary of non-valid categorical values

```{r}
dict_clean_categ <- check_categorical(
  ll1,
  dict_allowed = dict_categ1 # dictionary of categorical vars and their allowed values
)

dict_clean_categ
```

##### 2. Manually review non-valid values and give appropriate replacements, or use keyword ".na" to indicate that the value has been reviewed and cannot be corrected.

Again, we would normally do this step in a spreadsheet but we do it in R here for simplicity.

```{r}
dict_clean_categ$replacement <- c(
  "Years",
  "Years",
  "Cured",
  ".na",
  "M",
  ".na",
  "Suspected"
)
```

##### 3. Use the cleaning dictionary to clean categorical variables

```{r}
clean_categorical(
  ll1,
  dict_allowed = dict_categ1,
  dict_clean = dict_clean_categ
)
```

##### 4. If the original dataset is updated, we can repeat the cleaning steps while retaining the corrections already made in the previous cleaning dictionary, as in the example with numeric variables above.
